---
title: "Bike Sharing Data Analysis Project"
author: "Virginia Phan"
date: "2022-08-07"
output: html_document
---

## Import packages
```{r}
install.packages("tidyverse", repos = "https://cran.rstudio.com")
install.packages("skimr", repos = "https://cran.rstudio.com")
install.packages("tidyr", repos = "https://cran.rstudio.com")
install.packages("readr", repos = "https://cran.rstudio.com")
install.packages("ggplot2", repos = "https://cran.rstudio.com")
install.packages("dplyr", repos = "https://cran.rstudio.com")

```


```{r}
options(repos="https://cran.rstudio.com" )
```

## Import libraries
```{r}
library(tidyverse)
library(ggplot2)
library(tidyr)
library(lubridate)
library(readr)
library(plotly)
library(mapview)
library(dplyr)
library(scales)
```
## Import cleaned dataset
```{r}
trip_df <- read_csv('D:/Case Study 1 - Bike Sharing/Datasets/cleaned_tripdata.csv')
View(trip_df)
```
## Inspect dataframe
```{r}
View(trip_df)
str(trip_df)
colnames(trip_df)   # Columns's names of the dataframe
summary(trip_df)    # Statistical summary
tail(trip_df)
```
## DATA ANALYSIS
### Removing from dataframe where duration is invalid (<0)
```{r}
trip_df <- trip_df[!(trip_df$duration_in_second < 0),]
```

### Create tables number of rides for each types of memberships
#### Member's number of rides
```{r}
member_riders <- trip_df %>% 
                  filter(trip_df$membership_type == "member")

member_start <- member_riders %>% 
                  group_by(start_station_name, start_lat, start_lng) %>% 
                  summarise(n_rides_start = n()) %>% 
                  arrange(-n_rides_start)

member_start
```
#### Casual's number of rides
```{r}
casual_riders <- trip_df %>% 
  filter(trip_df$membership_type == "casual")

casual_start <- casual_riders %>% 
  group_by(start_station_name, start_lat, start_lng) %>% 
  summarise(n_rides_start = n()) %>% 
  arrange(-n_rides_start) 
casual_start
```

### Descriptive analysis on duration (in seconds)
```{r}
mean(trip_df$duration_in_second)     #(total duration/length)
max(trip_df$duration_in_second)      # longest duration
min(trip_df$duration_in_second)      # shortest duration
```
### Descriptive analysis on duration for each type of memberships
```{r}
avg_duration_mbs <- aggregate(trip_df$duration_in_second, by = list(trip_df$membership_type),FUN = mean)
avg_duration_mbs    # average duration for each type of memberships

max_duration_mbs <- aggregate(trip_df$duration_in_second, by = list(trip_df$membership_type), FUN = max)
max_duration_mbs    # longest duration for each type of memberships

min_duration_mbs <- aggregate(trip_df$duration_in_second, by = list(trip_df$membership_type), FUN = min)
min_duration_mbs    # shortest duration for each type of memberships


```
### Average duration by each day of week
```{r}
avg_duration_day <- aggregate(trip_df$duration_in_second, by = list(trip_df$day_of_week), FUN = mean)
avg_duration_day


# Reordering columns day_of_week
trip_df$day_of_week <- ordered(trip_df$day_of_week, levels = c('Monday', 'Tuesday', 'Wednesday', 'Thursday',
                                                               'Friday', 'Saturday', 'Sunday'))
```

### Ridership by each day of week for members and casuals
```{r}
ridership_day_mbs <- trip_df %>% 
                      group_by(day_of_week, membership_type) %>% 
                      summarise(num_of_ride = n(),
                                avg_duration = mean(duration_in_second)) %>% 
                      arrange(day_of_week, avg_duration, num_of_ride)
ridership_day_mbs
```
### Number of rides for each month of the year
```{r}
num_ride_per_month <- trip_df %>% 
  group_by(month, membership_type) %>% 
  summarise(number = n()) %>% 
  arrange(month, number)
num_ride_per_month


# Reorder column month
trip_df$month <- ordered(trip_df$month, levels = c('January', 'February', 'March', 'April', 'May', 'June', 'July',
                                                   'August', 'September', 'October', 'November', 'December'))
```
### Number of rides for each type of bike
```{r} 
num_of_ride_type <- trip_df %>% 
                      group_by(rideable_type) %>% 
                      summarise(number = n()) %>% 
                      arrange(-number)
num_of_ride_type
  
```

## DATA VISUALIZATION
### Number of rides each day of week
```{r}
count_per_day <- trip_df %>% 
                    group_by (day_of_week) %>% 
                    summarise(number = n())
count_per_day$day_of_week <- factor(count_per_day$day_of_week, levels = c("Monday", "Tuesday", 
                                                                          "Wednesday", "Thursday",
                                                                          "Friday", "Saturday",
                                                                          "Sunday"))
View(count_per_day)

ggplot(data = count_per_day, aes(x = day_of_week, y = number, fill = day_of_week))  + geom_bar(stat = "identity",
                                                                                               color = "black") + 
  geom_text(aes(label = number), vjust = 1.6, color = "black", size = 3.0) + labs(title = 'Number of Rides Each Day of the Week', y = 'Number of rides', x = 'Day of week') +
              scale_y_continuous(labels = comma) + theme_classic() 
```

### Casual vs Member uses of ride sharing throughout the day
```{r}
count_per_hour <- trip_df %>% 
                  group_by(time_of_day, membership_type) %>% 
                  summarise(number = n())
View(count_per_hour)

ggplot(data = count_per_hour, aes(x = time_of_day, y = number, group = membership_type)) + geom_line(aes(color = membership_type)) +
  geom_point(aes(color = membership_type, shape = membership_type)) + scale_x_continuous(breaks = c(0,1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23)) +
  scale_color_manual(values = c("#272324", "#E86F68")) + theme_minimal() +
  labs(title = "Casual vs Member uses of ride sharing throughout the day", x = 'Daytime', y = "Number of rides")

```

### Most prefered type of bikes (based on numbers of rides per year)
```{r}
proportion_bike_type <- trip_df %>% 
                        group_by(rideable_type)  %>% 
                        summarise(number = n()) %>% 
                        mutate(percent = round(number/sum(number)*100))

View(proportion_bike_type)


pie(proportion_bike_type$percent, labels = paste(proportion_bike_type$rideable_type, 
    sep = " ",proportion_bike_type$percent, "%" ),
    col = c('#82E0AA', '#AED6F1', '#F9E79F'),
    main = "Most prefered type of bikes (based on numbers of rides per year)")

```
### Average duration (in seconds) for each type of memberships
```{r}
average_length_membership <- trip_df %>% 
                              group_by(membership_type, rideable_type) %>% 
                              summarise(avg_ride_length = round(mean(duration_in_second)))



View(average_length_membership)



ggplot(data = average_length_membership, aes(x = membership_type, y = avg_ride_length, fill = rideable_type)) + 
  geom_bar(stat = "identity", position = "dodge") + 
  geom_text(aes(label = paste(avg_ride_length,"s")), vjust = 1.5, color = "black", size = 3.0,
                                        position = position_dodge(.9)) +
  scale_fill_brewer(palette = "Pastel1") + theme_light() +
  labs(title = "Average duration (in seconds) for each type of memberships", y = 'Average duration in seconds', x = 'Membership types')

```
### Number of rides by membership types throughout the year
```{r}
num_ride_per_month <- trip_df %>% 
                      group_by(month, membership_type) %>% 
                      summarise(number = n())
View(num_ride_per_month)


ggplot(data = num_ride_per_month,aes( x = month, y = number, group = membership_type)) +
  geom_line(aes(color = membership_type)) + geom_point(aes(color = membership_type, shape = membership_type)) +
  scale_x_discrete("Month", labels = c("Jan", "Feb","Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")) +
  scale_y_continuous(labels = comma) + theme_light() + 
  labs(title = 'Number of rides by membership types throughout the year', y = 'Number of rides')

```
### Top 5 prefered start station based on number of rides

```{r}
top_5_station <- trip_df %>% 
                  group_by(start_station_name) %>% 
                  summarise(number = n()) %>% 
                  arrange(desc(number)) %>% 
                  slice(1:5)

View(top_5_station)

ggplot(data = top_5_station, aes(x = reorder(start_station_name, number), y = number, fill = number)) + geom_bar(stat = "identity") +
  coord_flip() + labs(title = 'Top 5 prefered start station based on number of rides', x = 'Number of rides', y = 'Start station name') 

```

### Member Riders' Favourite Start Locations
```{r}
member_riders <- trip_df %>%
  filter(trip_df$membership_type == "member")
member_start <- member_riders %>% 
  group_by(start_station_name, start_lat, start_lng) %>% 
  summarise(n_rides_start = n()) %>%
  arrange(-n_rides_start)


View(member_start)

member_start[1:50,] %>%
  mapview(
    xcol = "start_lng", 
    ycol = "start_lat",
    zcol = "n_rides_start",
    cex = "n_rides_start",
    at = c(10000, 20000, 30000, 40000, 50000, 60000),
    crs = 4326,
    grid = F, 
    legend = T,
    layer.name = "Member Riders' Favourite Start Locations")
```


### Casual Riders's Favorite Start Location
```{r}
casual_riders <- trip_df %>% 
  filter(trip_df$membership_type == "casual")

casual_start <- casual_riders %>% 
  group_by(start_station_name, start_lat, start_lng) %>% 
  summarise(n_rides_start = n()) %>% 
  arrange(-n_rides_start) 


View(casual_start)

casual_start[1:50,] %>% 
  mapview(
    xcol = "start_lng",
    ycol = "start_lat",
    zcol = "n_rides_start",
    cex = "n_rides_start",
    at = c(10000,20000,30000,40000,50000,60000),
    crs = 4326,
    grid = F,
    legend = T,
    layer.name = "Casual Riders's Favorite Start Location")
```




```{r}

options(repos="https://cran.rstudio.com" )
```












